---
title: "exe1"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(survival)
library(tidyverse)
```

# Exercise on the Nelson-Aalen estimator {.unnumbered}

In the period 1962-77 a total of $205$ patients with *malignant melanoma* (cancer of the skin) were operated at Odense University hospital in Denmark.
A number of covariates were recorded at operation, and the patients were followed up until death or censoring at the **end of the study at December 31, 1977**. 
We will study death from malignant melanoma considering death from other causes as censorings. (Source: Andersen, Borgan, Gill & Keiding, Springer, 1993.)

You may read the data into R by the command:
```{r, cache=TRUE}
melanoma = read.table("https://www.med.uio.no/imb/english/research/centres/ocbe/courses/imb9335/melanoma.txt",header=T) %>% 
  as_tibble()
```

The data are organized with one line for each of the 205 patients, and with the following variables in the seven columns:

- status: status (1=death from disease, 2=censored, 4=death from other cause)
- lifetime: life time from operation in years
- ulcer: ulceration (1=present, 2=absent)
- thickn: tumor thickness in mm
- sex: 1=female, 2=male
- age: age at operation in years
- grthick: grouped tumor thickness [1: 0-1 mm (i.e. below 2 mm), 2: 2-4 mm (i.e. at least 2 mm, but below 5 mm) , 3: 5+ mm (i.e. 5 mm or more)]

The following commands provide a **Nelson-Aalen plot** for female melanoma patients. (The survival-library has to be loaded.)

```{r, cache = TRUE}
surv.f = survival::survfit(Surv(lifetime,status==1) ~ 1, data = melanoma, 
  subset = (sex == 1), ctype = 1) # ctype = 1 means NA estimator
surv.m = survival::survfit(Surv(lifetime,status==1) ~ 1, data = melanoma, 
  subset = (sex == 2), ctype = 1)

plot(surv.f, fun="cumhaz", mark.time=FALSE, conf.type="plain", xlim=c(0,10), ylim=c(0,0.80), main="Females", xlab="Years since operation", ylab="Cumulative hazard")
plot(surv.m, fun="cumhaz", mark.time=FALSE, conf.type="plain", xlim=c(0,10), ylim=c(0,0.80), main="Males", xlab="Years since operation", ylab="Cumulative hazard")
```

a)  Perform the commands and interpret the Nelson-Aalen plot.

b)  Make a Nelson-Aalen plot for males and compare with the plot for females.

To plot the Nelson-Aalen estimates for both genders in the same figure, we may give the commands:

```{r, cache=TRUE}
surv.sex = survival::survfit(Surv(lifetime,status==1) ~ strata(sex), data=melanoma, ctype=1)

plot(surv.sex, fun="cumhaz", mark.time=FALSE, lty=1:2, xlim=c(0,10), ylim=c(0,0.80), main="Gender", xlab="Years since operation", ylab="Cumulative hazard")

legend("topleft", c("Females","Males"), lty=1:2)
```

c)  Perform the commands and inspect the plot.

- Straight slope => **constant hazard** independent of gender

d)  Make Nelson-Aalen plots for patients with ulceration present and absent and interpret the plots. (Ulceration is "present" if the surface of the tumor viewed in a microscope show signs of ulcers and "absent" otherwise.)

```{r, cache=TRUE}
surv.ulcer = survival::survfit(Surv(lifetime,status==1) ~ strata(ulcer), data=melanoma, ctype=1)

plot(surv.ulcer, fun="cumhaz", mark.time=FALSE, lty=1:2, xlim=c(0,10), ylim=c(0,0.80), main="Ulceration", xlab="Years since operation", ylab="Cumulative hazard")

legend("topleft", c("Present","Absent"), lty=1:2)
```
e)  Make Nelson-Aalen plots for the three thickness groups 0-1 mm, 2-4 mm, 5+ mm and interpret the plots.

```{r, cache=TRUE}
surv.grthick = survival::survfit(Surv(lifetime,status==1) ~ strata(grthick), data=melanoma, ctype=1)

plot(surv.grthick, fun="cumhaz", mark.time=FALSE, lty=1:3, xlim=c(0,10), 
  ylim=c(0,0.80), main="Tumor thickness", xlab="Years since operation", ylab="Cumulative hazard")

legend("topleft", c("0-1 mm","2-4 mm", "5+ mm"), lty=1:3)
```

# Exercise on Kaplan-Meier estimator and log-rank test

In this exercise, we will use the Kaplan-Meier estimator and the log-rank test to study survival for the melanoma patients.

We will consider Kaplan-Meier estimates for the mortality from malignant melanoma treating death from other causes as censoring.

We may compute and plot the Kaplan-Meier estimate of the survival distribution for male patients by the commands (you need to load the survival-library)

```{r}
fit.m=survfit(Surv(lifetime,status==1)~1,data=melanoma, subset=(sex==2), conf.type="plain")
plot(fit.m, mark.time=FALSE, xlab="Years after operation", main = "Males")
abline(h = 0.75, col="red")
abline(h = 0.5, col="blue")
```

To obtain a summary of the results, you may give the command:
```{r, cache=TRUE}
#?quantile.survfit
summary(fit.m)
```
a) Perform these commands and interpret the Kaplan-Meier plot. Determine the lower quartile of the survival distribution for males with 95% confidence limits using the output from the summary-command. (Note that the lower quartile corresponds to 75% survival probability.)

- Look at the **first time** the Survival probability falls below $75\%$

b) We may obtain the quartiles of the survival distribution for males by the command

```{r}
quantile(fit.m)
```

Perform this command and compare with the result you obtained in a).

c) Make a Kaplan-Meier plot for females, and determine the lower quartile for females with 95% confidence limits (if possible). Compare with the results for males.

```{r}
fit.f=survfit(Surv(lifetime,status==1)~1,data=melanoma, subset=(sex==1), conf.type="plain")
plot(fit.f, mark.time=FALSE, xlab="Years after operation", main = "Females")
quantile(fit.f)
```
d) Use the log-rank test to test the null hypothesis that males and females have the same mortality from malignant melanoma:

```{r}
survdiff(Surv(lifetime,status==1) ~ sex,data=melanoma)
```
What may you conclude from the test?

e) Make Kaplan-Meier plots for patients with ulceration present and absent and interpret the results. Is it possible to estimate the lower quartile for both ulceration groups? Estimate the lower quartile with confidence limits if possible. Is there a significant difference in cancer mortality for patients with ulceration present and absent?
```{r}
plot(surv.ulcer, mark.time=FALSE, lty = 1:2, xlab="Years after operation", main = "Ulceration")
legend("bottomleft", c("Present","Absent"), lty=1:2)
```

```{r}
quantile(surv.ulcer)
```

```{r}
# NO STRATA IN THE FORMULA NEEDED!
survdiff(formula = Surv(lifetime, status == 1) ~ ulcer, data = melanoma)
```

f) Make Kaplan-Meier plots for the three thickness groups 0-1 mm, 2-4 mm, 5+ mm and interpret the plots. Estimate the lower quartile with confidence limits if possible. Is there a significant difference in cancer mortality between the three thickness groups?

- Same procedure as above!!!
